# Genome-Wide Association Study {#gwas-testing}

```{r gwas_basics_association_testing, message = FALSE, warning = FALSE, echo = FALSE, eval = TRUE}
COURSE_loc = "~/Desktop/practical"
# COURSE_loc = "/Users/swvanderlaan/git/swvanderlaan/A_primer_in_Human_Cardiovascular_Genetics"

source(paste0(COURSE_loc, "/scripts/functions.R"))
source(paste0(COURSE_loc, "/scripts/colors.R"))
source(paste0(COURSE_loc, "/scripts/packages.R"))

```


Now that you have learned how to perform QC, you can easily run a GWAS and execute some downstream visualization and analyses. Let's do this with a dummy dataset.

## Exploring the data

Even though someone says that the QC was done, it is still wise and good practice to run some of the commands above to get a 'feeling' about the data. So let's do this.

```
plink --bfile gwas/gwa --freq --out gwas/gwa

plink --bfile gwas/gwa --missing --out gwas/gwa

plink --bfile gwas/gwa --hardy --out gwas/gwa
```

Let's visualize the results.

```{r load-data-gwas}
gwas_HWE <- data.table::fread(paste0(COURSE_loc, "/gwas/gwa.hwe"))
gwas_FRQ <- data.table::fread(paste0(COURSE_loc, "/gwas/gwa.frq"))
gwas_IMISS <- data.table::fread(paste0(COURSE_loc, "/gwas/gwa.imiss"))
gwas_LMISS <- data.table::fread(paste0(COURSE_loc, "/gwas/gwa.lmiss"))
```


```{r show-hwe-gwas, message=FALSE, warning=FALSE}
gwas_HWE$logP <- -log10(gwas_HWE$P)

ggpubr::gghistogram(gwas_HWE, x = "logP",
                    add = "mean",
                    add.params = list(color = "#595A5C", linetype = "dashed", size = 1),
                    rug = TRUE,
                    # color = "#1290D9", fill = "#1290D9",
                    color = "TEST", fill = "TEST",
                    palette = "lancet",
                    facet.by = "TEST",
                    bins = 50,
                    xlab = "HWE -log10(P)") +
  geom_vline(xintercept = 5, linetype = "dashed",
                color = "#E55738", size = 1)
```


```{r show-freq-gwas, message=FALSE, warning=FALSE}
ggpubr::gghistogram(gwas_FRQ, x = "MAF",
                    add = "mean", add.params = list(color = "#595A5C", linetype = "dashed", size = 1),
                    rug = TRUE,
                    color = "#1290D9", fill = "#1290D9",
                    xlab = "minor allele frequency") +
  geom_vline(xintercept = 0.05, linetype = "dashed",
                color = "#E55738", size = 1)
```


```{r show-sample-callrate-gwas, message=FALSE, warning=FALSE}
gwas_IMISS$callrate <- 1 - gwas_IMISS$F_MISS

ggpubr::gghistogram(gwas_IMISS, x = "callrate",
                    add = "mean", add.params = list(color = "#595A5C", linetype = "dashed", size = 1),
                    rug = TRUE, bins = 50,
                    color = "#1290D9", fill = "#1290D9",
                    xlab = "per sample call rate") +
  geom_vline(xintercept = 0.95, linetype = "dashed",
                color = "#E55738", size = 1)
```


```{r show-snp-callrate-gwas, message=FALSE, warning=FALSE}
gwas_LMISS$callrate <- 1 - gwas_LMISS$F_MISS

ggpubr::gghistogram(gwas_LMISS, x = "callrate",
                    add = "mean", add.params = list(color = "#595A5C", linetype = "dashed", size = 1),
                    rug = TRUE, bins = 50,
                    color = "#1290D9", fill = "#1290D9",
                    xlab = "per SNP call rate") +
  geom_vline(xintercept = 0.95, linetype = "dashed",
                color = "#E55738", size = 1)
```


## Genetic models
A simple chi-square test of association can be done.

```
plink --bfile gwas/gwa --model --out gwas/data
```

_Genotypic_, _dominant_ and _recessive_ tests will not be conducted if any one of the cells in the table of case-control by genotype counts contains less than five observations. This is because the chi-square approximation may not be reliable when cell counts are small. For SNPs with MAFs < 5%, a sample of more than 2,000 cases and controls would be required to meet this threshold and more than 50,000 would be required for SNPs with MAF < 1%.

You can change this default behaviour by adding the flag `--cell`, _e.g._, we could lower the threshold to 3.

```
plink --bfile gwas/gwa --model --cell 3 --out gwas/data
```

Let's review the contents of the results.

```{r load-gwas-model}
gwas_model <- data.table::fread(paste0(COURSE_loc, "/gwas/data.model"))

dim(gwas_model)

N_SNPS = length(gwas_model$SNP)

gwas_model[1:10, 1:10]
```

It contains `r N_SNPS` rows, one for each SNP, and each type of test (_genotypic_, _trend_, _allelic_, _dominant_, and _recessive_) and the following columns:

- chromosome [CHR],
- the SNP identifier [SNP],
- the minor allele [A1] (PLINK always codes the A1-allele as the minor allele!),
- the major allele [A2],
- the test performed [TEST]:
  - GENO (genotypic association);
  - TREND (Cochran-Armitage trend);
  - ALLELIC (allelic as- sociation);
  - DOM (dominant model); and
  - REC (recessive model)],
- the cell frequency counts for cases [AFF], and
- the cell frequency counts for controls [UNAFF],
- the chi-square test statistic [CHISQ],
- the degrees of freedom for the test [DF],
- and the asymptotic P value [P] of association.

## Logistic regression
We can also perform a test of association using logistic regression. In this case we might want to correct for covariates/confounding factors, for example age, sex, ancestral background, i.e. principal components, and other study specific covariates (e.g. hospital of inclusion, genotyping centre etc.). In that case each of these P values is adjusted for the effect of the covariates.

When running a regression analysis, be it linear or logistic, PLINK assumes a multiplicative model. By default, when at least one male and one female is present, sex (male = 1, female = 0) is automatically added as a covariate on X chromosome SNPs, and nowhere else. The `sex` flag causes it to be added everywhere, while `no-x-sex` excludes it.

```
plink --bfile gwas/gwa --logistic sex --covar gwas/gwa.covar --out gwas/data
```

Let's examine the results
```{r load-gwas-assoc}
gwas_assoc <- data.table::fread(paste0(COURSE_loc, "/gwas/data.assoc.logistic"))

dim(gwas_assoc)

gwas_assoc[1:9, 1:9]
```

If no model option is specified, the first row for each SNP corresponds to results for a multiplicative test of association. The C >= 0 subsequent rows for each SNP correspond to separate tests of significance for each of the C covariates included in the regression model. We can remove the covariate-specific lines from the main report by adding the `hide-covar` flag.

The columns in the association results are:
- the chromosome [CHR],
- the SNP identifier [SNP],
- the base-pair location [BP],
- the minor allele [A1],
- the test performed [TEST]: ADD (multiplicative model or genotypic model testing additivity),
  - GENO_2DF (genotypic model),
  - DOMDEV (genotypic model testing deviation from additivity),
  - DOM (dominant model), or
  - REC (recessive model)],
- the number of missing individuals included [NMISS],
- the OR relative to the A1, _i.e._ minor allele,
- the coefficient z-statistic [STAT], and
- the asymptotic P-value [P] of association.

We need to calculate the standard error and confidence interval from the z-statistic. We can modify the effect size (OR) to output the beta by adding the `beta` flag.


## Let's get visual

Looking at numbers is important, but it won't give you a perfect overview. We should turn to visualizing our results in Chapter \@ref(gwas-visuals).
