# Phenome-Wide Association Study (PheWAS)

```{r phewas, message = FALSE, warning = FALSE, echo = FALSE, eval = TRUE}
COURSE_loc = "~/Desktop/practical"
# COURSE_loc = "/Users/swvanderlaan/git/swvanderlaan/A_primer_in_Human_Cardiovascular_Genetics"

source(paste0(COURSE_loc, "/scripts/functions.R"))
source(paste0(COURSE_loc, "/scripts/colors.R"))
source(paste0(COURSE_loc, "/scripts/packages.R"))

```

Given the large biobanks available nowadays that have also genotyped the participants and collected a vast-array of information on them, it is possible to perform a phenome-wide association study, a PheWAS.

There are several options in (Table \@ref(tab:tab-phewas)).

```{r, eval = TRUE, echo = FALSE}
phewas <- openxlsx::read.xlsx("information.xlsx", sheet = "PheWAS")
```


```{r tab-phewas, tidy = FALSE, eval = TRUE, echo = FALSE}
knitr::kable(
  phewas, 
  caption = 'PheWAS Resources.',
  # align = ,
  booktabs = FALSE
)
```


*Assignment*

1. Perform a PheWAS with a few of the resources and your favorite SNP from this tutorial. 
2. Compare the different websites. What do you notice?
3. How is a PheWAS informative?


## What's next?

You now know to what other traits, risk factors and diseases our trait of interest (coronary artery disease) is correlated. An obvious next question would be: is this a causal correlation? We will cover this aspect in the next Chapter \@ref(mendelian_randomization). 
